\chapter{Background \& Literature Overview}

%In this section you need to explain all the theory required to understand your dissertation (i.e.\ the following chapters). But really in this chapter I am going to show you some examples.

%Up to now, I have been using the term Christian games...

\section{What Are Christian Digital Games?} \label{sec:what_are_christian_digital_games?}

What makes something, or even someone, ``Christian''? Regardless of the subject of discussion, this question is always likely to spark conflicting views. For instance, \textcite{schultze_getting_2008} noted that fans of U2’s Bono have questioned his Christian status after he swore during an award show. The authors further observed that Christian groups also disagree on what it means for a cultural artefact to be Christian. For example, they mention that many Evangelicals consider media as Christian only if it is made by and for Evangelicals. Others, such as \textcite{lawton_what_2021}, \textcite{macinnis_what_2025}, or \textcite{unthank_what_2023}, may prefer to consider content, authorial intent, or audience impact when determining whether literature, music, or film is Christian.

Similarly, games scholar Kevin Schut states that currently there is no consensus on what defines a Christian game \parencite[8:38]{faith_forms_problem_2024}. This becomes apparent when one examines online discussion around the topic. For example, one of the factors that \textcite{moon_channel_why_2023} considers when determining if a digital game is Christian is that it must promote Christian morals. Meanwhile, \textcite{hartgrove_why_2022} considers \textit{Hypnospace Outlaw} (2019) a Christian game primarily because its creator, Jay Tholen, identifies as a Christian\footnote{Although Hartgrove does discuss the game's content somewhat, he does not deem explicit ideological content as necessary for a digital game to be Christian.}.

From the examples mentioned so far, it becomes evident that the question of what constitutes something as Christian is a highly contentious and subjective matter. Therefore, before analysing the opinions surrounding them, I needed to establish an operational definition of ``Christian digital games''.

\subsection{Existing Definitions}

Earlier work by \textcite{schut_making_2013} gives us a good starting point for such a definition. Based on a series of interviews with seventeen Christians working in the game industry, he outlined four definitions of ``Christian Games''. Explicitly Christian Games ``feature Bible stories or passages, very clear presentation of the gospel message, and stories that very openly have a Christian worldview''. As an example, Schut raises \textit{Big Bible Town} (2010), a digital game which aims to teach children about Bible stories. Meanwhile, Allegorical Christian Games ``have stories that are not necessarily openly Christian but have Christian themes or messages underneath them'' (p.~137). Interviewees mentioned no instances of such games, but they cited the Narnia series by C.S. Lewis as an example of this style.

Interviewees also raised what Schut terms as Evangelical Christian digital games, which ``focus on sharing the gospel with people who aren\textquotesingle t Christians''. (pp.~137--138). However, this is only a hypothetical definition since Schut could not cite any examples. Lastly, Christian-friendly games, such as Bejeweled and Words With Friends, are ``clean and safe, without significant violence, adult sexuality, swearing, anti-Christian worldviews, and other things that Christians might find offensive or disturbing'' (p. 137).

Schut\textquotesingle s work also gives valuable insight into how the term ``Christian Games'' is used. For example, he notes that most of the interviewees commonly defined ``Christian Games'' as either Explicitly Christian Games or Allegorically Christian games. However, only few of them considered Christian-friendly games part of the genre. Although this information proved useful for my study, I was hesitant to reuse these definitions due to Schut\textquotesingle s ambiguous use of the term ``Christian''; as I have already mentioned in the beginning of this section (see Section~\ref{sec:what_are_christian_digital_games?}), individuals may not necessarily agree on what determines something as Christian.

Subsequent work by \textcite{schut_swords_2026} acknowledges the subjective application of this term (p. 310), and clarifies his use of the term ``Christian''. While there are various groups which could be described as Christian, Schut observes that the Christian media industry in North America is influenced primarily by Evangelicals. He attributes their dominance to their collective focus on proselytism and their eagerness to adopt new media technologies as a means to that end. However, he also remarks that Evangelicals have not always counted for the majority of North American Christians, and that they are still garnering a presence in most other countries. Furthermore, he also notes that other groups, such as Roman Catholics, have also produced Christian games. Consequently, he specifies that, within the context of Christian digital games, the term ``Christian'' is often synonymous with ``Evangelical'', though this is not always the case.

Within the same study, Schut continued to investigate the ways in which Christians have employed the term ``Christian genre''. He found that, occasionally, Christians have argued that digital games with Christian authorship are Christian games. However, after a brief survey of the types of digital games which players have described as Christian, Schut concluded that Christians typically use the term to disclose the presence of Christian ideology within a digital game. However, Schut does not provide a more detailed, formal definition which I could use for my research.

Around 12 years earlier, \textcite{gonzalez_born-again_2014} had already shown that the overwhelming majority of Christian digital games had Evangelical associations by compiling a comprehensive list of religious digital games. However, unlike Schut, Gonzales does not opt to define Christian games for her research. Instead, she defines ``religious games'' to study Evangelical digital games within the broader picture of playful religious software. Using ``agential realism'' as an ontological frame, Gonzales describes something as ``religious'' if it stages interactions between human and non-human beings (p. 24), such as angels and demons.

Based on this definition, she excludes games like \textit{Castlevania} (1986) from her study, which include typically religious materials to embellish their fictional world. She does this because these games do not depict the non-human others as agents with which borders must be negotiated, but as imaginary entities. Similarly, she disregards games such as \textit{Run, Jesus Run!} (2010) or \textit{Crucify-me! Jesus} (2000), which utilize religious material derisively, since these negotiate boundaries between human groups. Furthermore, Gonzales limits her study to religious organizations, which she defines as a long-lived alliance of human and non-human entities which jointly determine the borders of humanity. 

While this definition is more rigorous than either of Schut's definitions, my original research question does not concern games associated with religions unrelated to Christianity. Therefore, I could not use it for my research. Nevertheless, her definition of religious organisations proved useful as a source of inspiration for my own definition, which I describe in the following section.

\subsection{My Definition} 

Each of the definitions mentioned in the previous section presents valid ideas about what can qualify a game as ``Christian'' or ``religious''. However, given that the definitions are either incomplete or cover games outside my field of study, I could not reuse them for my research. Therefore, using these existing definitions as a point of departure, I created my own definition of ``Christian digital games'' (or ``Christian games'' for short).

As I have already pointed out in the previous section, a complete definition of Christian games also needs to elucidate its use of the term ``Christian''. During my research, I could not find much insight into how players or developers use the term in the context of Christian games. While I have found blogs by \textcite{oliver_why_2014, oliver_christian_2016} which mention Evangelicals when discussing Christian games, there are also other individuals which do not specify any particular denomination \parencite{hartgrove_why_2022, moon_channel_why_2023, monkastery_why_2018}\footnote{Nevertheless, it is worth noting that the examples cited within these discussions are Evangelical games.}. Meanwhile, related studies either fail to elaborate on their usage \parencite{schut_making_2013} or employ it as a catch-all term for various religious groups \parencite{bernauer_playing_2012, schut_swords_2026}. Furthermore, within her list of religious games, \textcite{gonzalez_born-again_2014} seems to use ``Christian'' as an overarching category for games associated with Christianity.

Therefore, to facilitate discussion about existing research, I defined ``Christianity'' as blanket term encompassing a constellation of long-lived groups or institutions which promote their interpretation of Christ\textquotesingle s teachings, such as Evangelicals, Presbyterians, Anglicans, Roman Catholics, Greek Orthodox, and Mormons. Similarly, something (e.g. an ideology or a belief) is ``Christian'' when it is primarily based on the teachings of Jesus Christ as understood by one or more of these communities. Of course, these definitions are by no means authoritative; they only serve to establish a common understanding for discussion.

After establishing a working definition of ``Christian'', I proceeded by defining Christian games as a genre. Given that I was investigating players\textquotesingle\ perceptions of Christian digital games, I based my definition on the most common uses of the term, which relate to Christian ideology \parencite{schut_swords_2026}. Other elements, such as biblical stories, Bible quotes, Christian themes and messages, which \textcite{schut_making_2013} outlines in his definitions of Explicitly Christian games and Allegorical Christian games, could be said to be manifestations of such a worldview. Additionally, \textcite{gonzalez_born-again_2014} remarks that many Evangelical games betray a theology of spiritual warfare prevalent among Evangelicals in the 1980s and 90s, even if ideology is not a prerequisite in her definition of religious games.

Therefore, for this study, ``Christian games'' are digital games with an underlying Christian worldview that presents itself through legible signifiers. That is, the game contains elements which players can trace to Christian ideology with relative ease. Obvious examples of such signifiers include adaptations of Bible stories and the inclusion of Bible quotes, such as in \textit{Super 3D Noah\textquotesingle s Ark} (1994). Other Christian games indicate their Christian status through their treatment of their scripted narrative\textquotesingle s themes. Consider \textit{That Dragon, Cancer} (2016) for example. Its scripted narrative tackles the loss of the developers\textquotesingle\  son through the lens of their Christian faith, making their spirituality explicit in the process.

Alternatively, scripted narratives in Christian games can feature thinly veiled Christian allegories. The developers of \textit{Alum} (2015), describe it as ``a sort of Christian Allegory along the lines of Pilgrims [sic] Progress or The Lion, the Witch, and the Wardrobe'' \parencite{crashable_studios_alum_2015}. Some players recognised this, calling Alum a Christian game, often citing, or implying, its narrative as their primary reason\footnote{For examples, see \textcite{arkane_review_2015}, \textcite{hamja_review_2015} and \textcite{virotti_review_2017}.}. One player specifies that they classify it as a Christian game because the game\textquotesingle s scripted narrative develops a loving relationship between the eponymous player character, ``Alum'', and a deity called the ``Unfeigned Altruist'', while also pitting Alum in a fight against dark forces\footnote{See Chapter~\ref{ch:analysis_and_discussion} for a more detailed analysis of \textit{Alum}\textquotesingle s narrative.} \parencite{thomascanarias_review_2016}. Even if players do not notice the parallels to Christianity while playing, the game\textquotesingle s ending makes one parallel explicit when it calls Jesus Christ ``the true and living Unfeigned Altruist'' \parencite[10:20]{innuendo_studios_co-vids_2020}.

For the most part, \textit{Alum} (2015) attempts to be more covert than \textit{Super 3D Noah\textquotesingle s Ark} (1994) or \textit{That Dragon, Cancer} (2016) about its ties to Christianity. Nevertheless, all three narrow the range of possible associations such that even players with a cursory knowledge of Christianity can make the connection. The same cannot be said about the example given by \textcite{hartgrove_why_2022}. He argues that through its aesthetically pleasing qualities, artistic excellence, and believable characters, \textit{Hypnospace Outlaw} (2019) reveals its creator\textquotesingle s (i.e. Jay Tholen\textquotesingle s) love for beauty originating from his Christian beliefs. While this philosophy can be connected to Christianity, it is hardly exclusive to it. Without prior knowledge of Tholen\textquotesingle s religious beliefs, players may just as easily relate it a secular worldview.

Since my definition requires the digital game in question to be aligned with Christian ideology, I did not consider games which draw on Christianity for inspiration. So, despite requiring players to journey through hell and slaughter demon hordes, \textit{Doom} (1993) falls outside of the field being studied. The same cannot be said for games which reference Christianity to mock it or its followers. As an example, \textcite[120]{gonzalez_born-again_2014} mentions BoMToons, a Mormon game developer which created multiple digital games poking fun at Mormon scripture and culture. However, the examples brought up by players when criticising Christian games, such as \textit{Left Behind: Eternal Forces} (2006) \parencite{hartgrove_why_2022} and \textit{Captain Bible in Dome of Darkness} (1994) \parencite{moon_channel_why_2023}, tend to be more earnest about their subject matter. Hence, I limited my study to digital games which approach Christianity in a more solemn manner.

\section{Related Work}
%\textbf{In this section you need to explain (and reference) similar work in literature}.  Make sure to:

\if{false}
  \begin{itemize}
   \item Give a systematic overview of papers with related/similar work
   \item Highlight similarities/differences to your work (perhaps in the form of a table)
  \end{itemize}

  Note that this section may be sectioned based on the different aspects of your dissertation.
\fi

With my area of study established, I am now in a position to discuss related research examining mainstream reception of Christian games. As I will demonstrate shortly, this study aims to address an existing research gap by examining the content of Christian games as a potential cause of these critiques.

%TODO Explain more about Schut (2013), it mentions the opinions under study, but does not delve into why they hold them.

%TODO Is game makers even a term?

In the introduction, I have already mentioned the interviews by \textcite{schut_making_2013}, which discuss the opinions of 17 Christian game-makers regarding Christian games. Throughout my research, I was unable to find another study analysing the opinions surrounding Christian games. \textcite{gonzalez_born-again_2014} briefly mentions bloggers of unnoticed games mocking Christian games (p. 8), and claims of unoriginality (pp. 143--144), an issue closely related to the one under examination. However, her dissertation mostly avoids discussing the enjoyment derived from Christian games. When reflecting on her own experiences playing Evangelical games, she recalls occasions of bewilderment only to dismiss them. Instead, she asserts that she found Evangelical games to be ``wonderful, unforgettably good games'' (p. 266). Nevertheless, these moments of discomfort may merit further attention; understanding their cause might help clarify the reasons behind certain player\textquotesingle s dismay, and help future developers to avoid these pitfalls. While it may be tempting to dismiss all criticism as snide remarks, it is beneficial to give an honest consideration to negative feedback. This is doubly true when that feedback comes from one\textquotesingle s target audience.

Although neither \textcite{bernauer_playing_2012} or \textcite{schut_swords_2026} mention the negative opinions surrounding Christian games, both propose reasons for their lack of success. For one, Bernauer maintains that the lack of mainstream interest in Christian games is, in part, due to the explicit Christian convictions presented within these games. Furthermore, she also blames Christian games\textquotesingle\ (mostly) young target audience, since studies found the average gamer to be around thirty-five years old. Lastly, she presents several issues stemming from a lack of finances when comparing Christian game companies to other studios. These include their inability to publish their games to consoles, the lack of both local and online multiplayer in a majority of their titles, and the short lifespan of these firms.

Meanwhile, Schut attributes three reasons for the lack of commercial success of Christian games within Evangelical circles. Primarily, he ascribes this predicament to the lack of blockbusters within the Christian games industry; unlike Christian television, for instance, Christian games do not have their own \textit{VeggieTales} to drive sales. He also notes that the moral controversies surrounding \textit{Dungeons \& Dragons} in the 1980s may soured Evangelicals\textquotesingle\ opinions around Christian games. Additionally, he cites an argument by \textcite{moon_channel_why_2023} regarding the Protestant work ethic within America, which conflicts with thriftless, unproductive play. However, Schut admits that Evangelical\textquotesingle s receptiveness towards consumerism discredits this explanation somewhat.

Although all of these reasons may indeed contribute to the unpopularity of Christian games, most of them are factors external to the games\textquotesingle\ development process. Bernauer offers actionable suggestions for some of the issues raised, such as targeting older age demographics and making Christian content more implicit. However, most of the reasons given by both researchers are difficult or even impossible for Christian game developers to control. Furthermore, both of them also fail to acknowledge the main reason why many players have decried Christian games in the first place --- the games\textquotesingle\ content. Hence, in the following section I examine two Christian games in the hopes of better understanding what causes certain players to view Christian games as preachy.


\section{Summary}
%\blindtext
